---
title: "Species Distribution Modeling"
author: "Ridwan Shittu"
date: "2023-02-20"
output: 
  html_document:
    toc: true # table of content true
    toc_depth: 3
    toc_float: true
    theme: united 
---

#Species Distribution Modelling

## Settings
### Load required libraries and create folders
```{r library, message=FALSE, warning=FALSE}
#load R-Packages
library(raster) #for Raster analysis
library(ENMeval) #for testing different models and feature-for model selection 
library(dismo) #for using Maxent in R
library(sf) 
library(dplyr)
library(geodata) #to download worldclim new version dataset
library(rnaturalearth) # for getting different shapefiles
library(rJava) # require to run Maxent in R.
#NB. One can use Maxnet also in R
```

## Working directory and folders
Set working directory and create folder for data
```{r directoryfolder, message=FALSE, warning=FALSE}
#Create folders
getwd() # determine your working directory

#Check if data folder is on the working directory,if not create data folder to hold the process data
if(!file.exists("occ")){
  dir.create("occ")
}
if(!file.exists("env")){dir.create("env")}
if(!file.exists("env/world")){dir.create("env/world")}
if(!file.exists("env/study_bg")){dir.create("env/study_bg")}
if(!file.exists("env/europe")){dir.create("env/europe")}
```

## Clean occurrence records
load or download occurence records of species or phenomenol of interest.
Clean and process the occurrence and environmental data to remove duplicates, errors, or missing values.
Here we will work with the  occurrence records of Aedes albopictus available as supplementary data for the publication of Kraemer et al 2015: 
```{r occurrence data, message=FALSE, warning=FALSE}
## load occurrence records
alb_raw <- read.csv("occ/albopictus_corrected.csv", na.strings = "NA", stringsAsFactors = FALSE)
head(alb_raw) #check the data and identify the coordinate columns
colnames(alb_raw)[c(6,7)] <- c("lat", "lon")
colnames(alb_raw)
#clean the occurrence records
#check for NAs and duplicates
sum(is.na(alb_raw[,6:7]))==0 # check for coordinates with NAs
#remove nas
alb_clean <- subset(alb_raw, !is.na(lon) & !is.na(lat))
head(alb_raw)
cat(nrow(alb_raw)-nrow(alb_clean), "records are removed")

sum(duplicated(alb_raw[, 6:7]))==0 #check for duplicated records
#remove duplicated records
dups <- duplicated(alb_clean[c("lat","lon")])
alb_unique <- alb_clean[!dups,]
cat(nrow(alb_clean)-nrow(alb_unique), "records are removed")
dim(alb_raw);dim(alb_unique)
# inspect the duplicated records
str(alb_unique)
hist(as.numeric(alb_unique$YEAR), xlab="Year", main="Temporal distribution of Aedes albopictus")
```

# Environmental Data
## Download environmental data 
Download environmental data e.g. bioclimatic data.
Preprocess the environmental data to ensure they have the same resolution, projection, and spatial extent
```{r environmental data, message=FALSE, warning=FALSE}
#bio=bioclimatic variables from worldclim 2 (Fick &Hijman 2017). It has 19 layers of derived from monthly temperature and rainfall values in order to generate more biologically meaningful variables.
#bio
# res:2.5
bio <-  worldclim_global(var="bio", res=2.5, path="env/world/", version="2.1")
bios <- raster::stack(bio) #resaving spatRaster to raster

plot(bios[[1]])

# writeRaster(bios,filename=paste0(getwd(),"/",names(bios),".asc"),format="ascii", bylayer=TRUE,overwrite=T)
```


## Thin occurrence records
## Buffer occurrence
## Variable sel- corrélation matrix
## Variable selection-ENMeval
## Variable selection-Jackknife
## Partial Receiver operating curve
## Clamping and extrapolation
## Main Model
## Threshold
## Future Projection

