---
title: "Species Distribution Modeling"
author: "Ridwan Shittu"
date: "2023-02-20"
output: 
  html_document:
    toc: true # table of content true
    toc_depth: 2
    toc_float: true
    theme: united 
---

#Species Distribution Modelling
Species distribution modeling (SDM) is a statistical approach used to predict the potential distribution of a species based on environmental variables.
It can help in identifying suitable habitats for species or at risk of MBD transmission, guiding surveillance efforts, and predicting the impacts of climate change.
## Settings
### Load required libraries and create folders
```{r library, message=FALSE, warning=FALSE}
#load R-Packages
library(raster) #for Raster analysis
library(ENMeval) #for testing different models and feature-for model selection 
library(dismo) #for using Maxent in R
library(sf) 
library(dplyr)
library(geodata) #to download worldclim new version dataset
library(rnaturalearth) # for getting different shapefiles
library(rJava) # require to run Maxent in R.
#NB. One can use Maxnet also in R
source("prepPara.R") # function to tune maxent i.e. different from the autofeatures in maxent
```

## Working directory and folders
Set working directory and create folder for data
```{r directoryfolder, message=FALSE, warning=FALSE}
#Create folders
getwd() # determine your working directory

#Check if data folder is on the working directory,if not create data folder to hold the process data
if(!file.exists("occ")){
  dir.create("occ")
}
if(!file.exists("env")){dir.create("env")}
if(!file.exists("env/world")){dir.create("env/world")}
if(!file.exists("env/study_bg")){dir.create("env/study_bg")}
if(!file.exists("env/europe")){dir.create("env/europe")}
```

## Clean occurrence records
load or download occurence records of species or phenomenol of interest.
Clean and process the occurrence and environmental data to remove duplicates, errors, or missing values.
Here we will work with the  occurrence records of Aedes albopictus available as supplementary data for the publication of Kraemer et al 2015: 
```{r occurrence data, message=FALSE, warning=FALSE}
## load occurrence records
alb_raw <- read.csv("occ/albopictus_corrected.csv", na.strings = "NA", stringsAsFactors = FALSE)
head(alb_raw) #check the data and identify the coordinate columns
colnames(alb_raw)[c(6,7)] <- c("lat", "lon")
colnames(alb_raw)
#clean the occurrence records
#check for NAs and duplicates
sum(is.na(alb_raw[,6:7]))==0 # check for coordinates with NAs
#remove nas
alb_clean <- subset(alb_raw, !is.na(lon) & !is.na(lat))
head(alb_raw)
cat(nrow(alb_raw)-nrow(alb_clean), "records are removed")

sum(duplicated(alb_raw[, 6:7]))==0 #check for duplicated records
#remove duplicated records
dups <- duplicated(alb_clean[c("lat","lon")])
alb_unique <- alb_clean[!dups,]
cat(nrow(alb_clean)-nrow(alb_unique), "records are removed")
dim(alb_raw);dim(alb_unique)
# inspect the duplicated records
str(alb_unique)
hist(as.numeric(alb_unique$YEAR), xlab="Year", main="Temporal distribution of Aedes albopictus")

write.csv(alb_unique, "occ/albopictus_cleaned_1.csv")

#Inspect the occurrence data in QGIS for interactive editting.
```

# Environmental Data
## Download environmental data 
Download environmental data e.g. bioclimatic data.
Preprocess the environmental data to ensure they have the same resolution, projection, and spatial extent
```{r environmental data, message=FALSE, warning=FALSE}
#bio=bioclimatic variables from worldclim 2 (Fick &Hijman 2017). It has 19 layers of derived from monthly temperature and rainfall values in order to generate more biologically meaningful variables.
#bio
# res:2.5
if( !file.exists( paste0("env/world/wc2.1_2.5m/")   )){
  bio <-  worldclim_global(var="bio", res=2.5, path="~/Documents/ENM_2023/SDM-DiMoC-Workshop1/env/world/", version="2.1")
}
              
bio <-list.files("~/Documents/ENM_2023/SDM-DiMoC-Workshop1/env/world/wc2.1_2.5m", pattern=".tif",full.names=T)


bio <- raster::stack(bio) #resaving spatRaster to raster

plot(bio[[1]])

#to use maxent from the GUI interface.
# writeRaster(bio,filename=paste0(getwd(),"/","env/world/",names(bio),".asc"),format="ascii", bylayer=TRUE,overwrite=T)
save(bio, file ="~/Documents/ENM_2023/SDM-DiMoC-Workshop1/bio.RData")
```
## Thin occurrence records
Thin occurrence records, retain only one record per raster cell
```{r thin_occ, message=FALSE, warning=FALSE}
alb_occ <- alb_unique[,c("lon","lat")] #select the coordinates 
alb_cell <- raster::extract(bio[[1]], alb_occ, cellnumbers=TRUE) #extract the raster values for each point
sum(duplicated(alb_cell[,1]))==0  #check for duplicate i.e. two points on one raster cell
#remove duplicate
alb_celldups <- duplicated(alb_cell[,1]) # 1 represent the cell number column.
alb_thin <- alb_occ[!alb_celldups,]
cat(nrow(alb_occ)-nrow(alb_thin), "records are removed") # check the remaining occurrence point.

#Visualized the thinned records
plot(bio[[1]], main="Mean annual temperature")
points(alb_thin)

# another method = spthin packages, it takes longer time to run but allow a definition of the distance to use to thin the occurrence records.
```
## clip data to a study area Europe
Download and reshape Europe Shapefiles
```{r, clip data to europe, message=FALSE, warning=FALSE}

EU_spdf <- ne_countries(continent = "europe")
#exclude Russia
EU_spdf <- EU_spdf[EU_spdf$sovereignt !="Russia",]
sp::plot(EU_spdf)
#create an extent using the raster function extent to describe EU only: removing French Guana etc.
EU_ext <- raster::extent(c(xmin= -12, xmax=34, ymin=34, ymax=71))
#clip the EU shapefile to EU_extent using crop function
EU_shp <- crop(EU_spdf, EU_ext)
plot(EU_shp, col="blue")

alb_eu <- filter(alb_thin, lat>34 & lat<71, lon>-12&lon<34) #select occurrence in Europe

#Clip bioclim data to Europe
bio.eu <- raster::crop(bio, EU_shp)
bio.eu <- raster::mask(bio.eu, EU_shp)

plot(bio[[1]], main="Mean annual temperature") #plot the 
points(alb_eu, col="red")
plot(bio.eu[[1]], main="Mean annual temperature for Europe") #zoom to europe
points(alb_eu, col="red")

# writeRaster(bio.eu,filename=paste0(getwd(),"/",names(bio.eu),".asc"),format="ascii", bylayer=TRUE,overwrite=T)
```

## Buffer occurrence
Create buffer arround the occurrence records to reduce area for pseudo-absence.
### Create Buffer 
```{r buffer, message=FALSE,warning=FALSE}
#buffering occurrence records
#conver to simple feature
alb_eu_sf <- sf::st_as_sf(alb_eu, coords=c("lon", "lat"), crs=4326)

#3035-ETRS89-extended/LAEA EUrope
alb_eu_sf <- sf::st_transform(alb_eu_sf, crs = 3035) #easy to use

#here we use 1000km, 
# How can we determine the buffer distance to use for the occurrence data.
#we could create different maxent models using different buffersizes to determine the best buffersize to use.
alb_buf <- sf::st_buffer(alb_eu_sf, dist = 1000000) %>% 
  sf::st_union() %>% 
  sf::st_sf() %>% 
  sf::st_transform(crs = 4326)

plot(bio[[1]], main="Mean annual temperature")
points(alb_eu)
plot(alb_buf, border="red", lwd=3, add=TRUE)

### clip environmental data with the occurrence buffer
bio_buf <- raster::crop(bio, alb_buf)  # clip raster to the buffer area
bio_buf <- raster::mask(bio_buf, alb_buf) # mask out the clipped area
# 
plot(bio_buf[[1]], main="Mean Annual Temperature")
points(alb_eu)
plot(alb_buf, border="red", lwd=3, add=TRUE)
```
## Backgroup point selection
### Background point selection
```{r select background points, message=FALSE, warning=FALSE}
set.seed(0323) #to allow selection of the same random sampling point.
bg <- dismo::randomPoints(mask=bio.eu[[1]],n=10000,p=alb_eu, excludep = T)
bg <- as.data.frame(bg)
head(alb_eu)
head(bg)
colnames(bg) <- colnames(alb_eu)

write.csv(bg, "occ/background_seed0323.csv", row.names = FALSE)
write.csv(alb_eu, "occ/aedes_albopictus_thinned.csv", row.names = FALSE)
alb_eu.mx <- data.frame(species="species", alb_eu)# the format for maxent must have colume for species
write.csv(alb_eu.mx,"occ/aedes_albopictus_thinned_mx.csv", row.names = FALSE )

```

## Prepare Data for model fitting 
```{r}
# get the same random sample for training and testing
set.seed(0323)
# randomly select 70% for training
selected <- sample(1:nrow(alb_eu),nrow(alb_eu)*0.70)
alb_train <- alb_eu[selected,] # selection for model training
alb_test <- alb_eu[-selected,] # selection for model testing


p <- raster::extract(bio_buf,alb_train)
dim(p)
# out <- which(is.na(rowSums(p))) # if a record has missing data (NA) then the sum of its rows is also NA... find these (if any)
# if (length(out) > 0) {
#    p <- p[-out, ]
#   }

# extracting env conditions for background/pseudo-absence
a <- raster::extract(bio_buf,bg)
dim(a)
# out <- which(is.na(rowSums(a))) # if a record has
# 
# if (length(out) > 0) {
#    a <- a[-out, ]
#   }
dim(a)

# repeat the number 1 as many times as the number of rows in p,
# and repeat 0 for the rows of background points
pa <- c(rep(1,nrow(p)), rep(0,nrow(a)))

sdata <- as.data.frame(rbind(p,a))

```


## Variable sel- corrélation matrix
variable Selection involve choosing the most relevant environmental variables that may affect the species' distribution using statistical tests, expert knowledge, or ecological theory.
statistical approach involve testing for collinearity- a high correlation between variables. If affect understand and predicting the relationship between environmental variable and response variable. high correlation can be set to 0.7 (Dorman et al.2013)

```{r var_select1, message=FALSE, fig.width=8, fig.height=8}
library(corrplot)
cormat_env <- sdata
names(cormat_env) <- gsub(pattern="wc2.1_2.5m_", replace= "", x=names(cormat_env))
colinearity <- cor(cormat_env, use="complete.obs", method = "spearman")
# saveRDS(colinearity, "results/colinearity.Rds")
corrplot(colinearity, type = "upper", method = "number", col=COL2 ( "RdBu", 10))
# corrplot.mixed(colinearity)

# corrplot(colinearity, method = "square", addCoef.col = "grey", tl.pos = 'd', cl.pos = "r", type = "upper", col=COL2 ( "RdBu", 10))

```
```{r}
# statistical selection of variable using collinearity plot

colnames(sdata)
#svs-statistical variable selection
selectedVar <- c("wc2.1_2.5m_bio_1","wc2.1_2.5m_bio_4","wc2.1_2.5m_bio_7","wc2.1_2.5m_bio_8","wc2.1_2.5m_bio_12", "wc2.1_2.5m_bio_13", "wc2.1_2.5m_bio_15", "wc2.1_2.5m_bio_16","wc2.1_2.5m_bio_19")
mod_svs <- maxent(x=sdata[selectedVar],
                  p=pa,
                path=paste0(getwd(),"/results/maxent_prj_svs"),
                  args=prepPara(userfeatures="LQHPT",
                            betamultiplier=1))

plot(mod_svs)
pred.svs.buf <- predict(mod_svs, bio_buf)

plot(pred.svs.buf)
```

## Variable selection-ENMeval
ENMeval Kass etal., 2021 
```{r ENMeval, warning=FALSE}
# tune.args <- list(fc = c("LQ", "LQH", "LQP", "LQHP"), rm = seq(0.5, 4, 0.5))
# alb_enm <- ENMevaluate(alb_train, bio_buf, bg, algorithm = "maxent.jar",
#                        tune.args = tune.args, partitions = "block", parallel = TRUE, numCores = 8)
# takes 96 minutes 55.9 seconds to run.
# save(alb_enm, file = "results/alb_enm.RData")
load("results/alb_enm.RData")
# enm.test.mn <- ENMevaluate(alb_train, bio_buf, bg, algorithm = "maxnet", 
#                       tune.args = tune.args, partitions = "testing", 
#                       occs.testing = alb_test, parallel = TRUE, numCores = 4)
# 
# enm.test.mx <- ENMevaluate(alb_train, bio_buf, bg, algorithm = "maxent", 
#                       tune.args = tune.args, partitions = "testing", 
#                       occs.testing = alb_test, parallel = TRUE, numCores = 4)
# 

```
### Model Selection

```{r enmeval_selectn, message=FALSE, warning=FALSE}
# alb_res <- eval.results(enm.test.mn )
# opt.aicc <- alb_res %>% filter(delta.AICc == 0)
# opt.aicc
# 
# opt.seq <- alb_res %>% 
#   filter(or.10p == min(or.10p)) %>% 
#   filter(auc.val == max(auc.val))
# opt.seq

alb_res <- eval.results(alb_enm )
opt.aicc <- alb_res %>% filter(delta.AICc == 0)
opt.aicc

opt.seq <- alb_res %>%
  filter(or.10p.avg == min(or.10p.avg)) %>%
  filter(auc.val.avg == max(auc.val.avg))
opt.seq

```

## Variable selection-Jackknife
using the jackknife 

## Main Model

## Threshold

## Future Projection

## Reference
Hijmans, R. & Elith, J. (2021) Species distribution modeling.
https://rspatial.org/raster/sdm/

https://biodiversityinformatics.amnh.org/open_source/maxent/Maxent_tutorial_2021.pdf


Kass JM, Muscarella R, Galante PJ, Bohl CL, Pinilla-Buitrago GE, Boria RA, Soley-Guardia M, Anderson RP (2021). “ENMeval 2.0: Redesigned for customizable and reproducible modeling of species’ niches and distributions.” Methods in Ecology and Evolution, 12(9), 1602-1608. https://doi.org/10.1111/2041-210X.13628.

https://jamiemkass.github.io/ENMeval/articles/ENMeval-2.0-vignette.html#resources


Fick, S.E. and R.J. Hijmans, 2017. WorldClim 2: new 1km spatial resolution climate surfaces for global land areas. International Journal of Climatology 37 (12): 4302-4315.

Dormann, C.F., Elith, J., Bacher, S., Buchmann, C., Carl, G., Carré, G., Marquéz, J.R.G., Gruber, B., Lafourcade, B., Leitão, P.J., Münkemüller, T., McClean, C., Osborne, P.E., Reineking, B., Schröder, B., Skidmore, A.K., Zurell, D. and Lautenbach, S. (2013), Collinearity: a review of methods to deal with it and a simulation study evaluating their performance. Ecography, 36: 27-46. https://doi.org/10.1111/j.1600-0587.2012.07348.x
